\documentclass[anon]{nesy2026}
% NeSy 2026 (PMLR/jmlr). Double-blind: [anon]. amsmath, amssymb, natbib, graphicx, url, algorithm2e auto-loaded.
\usepackage{booktabs}
\newcommand{\WMC}{\ensuremath{\mathrm{WMC}}}
\newcommand{\tw}{\ensuremath{\mathrm{tw}}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\bC}[1]{\boldsymbol{#1}}

\title[Machine-checked modular WMC for NeSy]{Machine-checked, modular weighted model counting for neuro-symbolic inference over large ontologies}

\clearauthor{\Name{Anonymous Author(s)} \Email{anon@example.org}\\
 \addr Anonymous Institution}

\begin{document}
\maketitle

\begin{abstract}
\noindent Neuro-symbolic learning couples a neural perception model with symbolic reasoning under a
\emph{conditional-independence assumption}: the perception model emits independent per-concept
marginals $p_i$, and the symbolic layer treats them as an independent product. This assumption is
known to impede optimization, distort uncertainty, and admit reasoning shortcuts. Repairing it
exactly means conditioning the factorized distribution on the symbolic theory, whose target
marginals $\mu_i$ are ratios of weighted model counts ($\WMC$); exact $\WMC$ is $\#$P-hard and, on a
description-logic ontology with tens of thousands of axioms, far beyond exact reasoning. We give the
soft, message-passing closure used in practice the two properties a deployable inference layer needs:
a correctness characterization and tractability at ontology scale. First, we prove in machine-checked
Lean~4 (no \texttt{sorry}; core axioms $\subseteq\{\texttt{propext},\texttt{Classical.choice},\texttt{Quot.sound}\}$) that the
soft closure of the \emph{true-path theory} is loopy belief propagation, and is the \emph{exact}
$\WMC$ marginal on the tree fragment for \emph{all} priors $q_i$, with error confined to
reconvergences (a diamond yields exact $5/6$ versus soft-OR $7/10$). Second, the Gene Ontology
constraint graph has treewidth $9$/$31$/$272$ across its three namespaces, so monolithic exact
$\WMC$ is infeasible on the largest; yet cutting a ${<}8\%$ reconvergence core makes $92$--$100\%$ of
the ontology exactly computable by a modular construction, the bounded soft approximation confined to
the small core. Third, across the neuro-symbolic spectrum---DeepProbLog (exact compilation walls at
${\sim}400$ terms), Scallop (top-$k$ drift concentrated exactly at the reconvergences our theory
identifies), and A-NeSI/NeSyDM (collapse on coupled structure, $0.97\!\to\!0.23$ and
$0.96\!\to\!0.03$--$0.13$ from $N{=}1$ to $N{=}4$)---ours is uniquely both scalable and
accuracy-characterized. The layer is demonstrated on protein function prediction in a companion
paper.
\end{abstract}

\section{Introduction}
Neuro-symbolic (NeSy) learning combines the perceptual strength of neural networks with
the structure and verifiability of symbolic reasoning: a perception model maps inputs to
the truth values of symbolic concepts, and a symbolic layer reasons over them under a fixed
theory. Almost universally this coupling rests on a \emph{conditional-independence
assumption}---the perception model emits independent per-concept marginals $p_i\in[0,1]$
and the symbolic layer treats the joint as the product
$q(\bC y)=\prod_i p_i^{y_i}(1-p_i)^{1-y_i}$
\citep{manhaeve2018deepproblog,xu2018semanticloss,ahmed2022spl}---which makes the layer
differentiable and underwrites the semantic-loss and probabilistic-layer constructions much
of the field is built on \citep{xu2018semanticloss,ahmed2022spl,deraedt2020survey}.

The same assumption is now understood as a principal source of failure. Because the model
factorizes over concepts the theory couples, the product places mass on impossible
configurations and withholds it from theory-induced correlations. This \emph{impedes
optimization} (flattening or fragmenting the loss landscape), \emph{distorts uncertainty}
(the factorized posterior cannot represent a satisfied constraint's dependencies), and
admits \emph{reasoning shortcuts} (low loss through assignments that satisfy the theory but
not its semantics) \citep{vankrieken2024independence,marconato2023reasoningshortcuts,marconato2024bears}.
These pathologies are structural to the assumption and sharpen as the theory couples more
variables.

Repairing it has a clean meaning: condition $q$ on the symbolic theory $T$, yielding
$p_T(\bC y)\propto[\bC y\models T]\,q(\bC y)$, and read off the marginals
$\mu_i=p_T(y_i{=}1)=\WMC(T\wedge y_i;\bC p)/\WMC(T;\bC p)$, ratios of \emph{weighted model
counts} (WMC)---exactly the semantic-probabilistic-layer object
\citep{ahmed2022spl,xu2018semanticloss}. But exact WMC is $\#$P-hard \citep{chavira2008wmc},
its compilation cost $\Theta(m\,2^{\tw})$ in the constraint-graph treewidth $\tw$
\citep{darwiche2002kcmap,darwiche2011sdd,bodlaender1993treewidth}; on a real ontology with
tens of thousands of axioms this is out of reach. Here deployed NeSy systems retreat to
soft, message-passing relaxations---monotone closures, top-$k$ provenance---whose
correctness is essentially never characterized. The field thus faces a dichotomy: exact but
not scalable, or scalable but with no account of when, where, or how much they err.

This paper closes that gap for the \emph{true-path theory}---the propositional grounding of
an ontology's hierarchical (is-a/part-of) and definitional axioms, the theory governing
protein function prediction over the Gene Ontology (GO)
\citep{ashburner2000go,geneontology2023}. It is the EL\textsuperscript{++} normal form of
subsumption and conjunction-definition axioms \citep{baader2005elenvelope,kazakov2014elk},
and the soft closure deployed against it is the monotone fixpoint of the EL completion: each
concept aggregates a normalized soft-OR of its sufficient conditions, capped by the
true-path constraint. Our observation is that this closure is not a black box but loopy
belief propagation \citep{pearl1988probabilistic} on the constraint graph, and so can be
both \emph{characterized} for correctness and made \emph{tractable} at ontology scale. Three
contributions.

\begin{enumerate}\itemsep2pt
\item \textbf{Machine-checked exactness.}
We formalize the closure in Lean~4 \citep{demoura2021lean,mathlib2020} and prove it is
belief propagation: for a target with $k$ independent children under the true-path
constraint, the exact WMC marginal equals the soft-OR update for \emph{all} $k$ and
\emph{all} priors, and the closure converges unconditionally. The error is localized exactly
at \emph{reconvergences}: on a diamond the exact marginal is $5/6$ while the soft-OR gives
$7/10$. The development carries no \texttt{sorry}, concrete-development axioms
$\subseteq\{\texttt{propext},\texttt{Classical.choice},\texttt{Quot.sound}\}$; a
kernel-checked battery confirms tree-exactness over an $81$-prior grid and diamond
inexactness across priors (\S\ref{sec:lean}).

\item \textbf{Modular exact WMC at ontology scale.}
We measure the GO treewidth---$9$/$31$/$272$ across cellular component/molecular
function/biological process---so monolithic exact WMC is feasible only on the smallest
($7.5$\,s; the soft closure agrees to mean error $6\times10^{-4}$, $7\times$ larger at
multi-parent nodes, the predicted reconvergence signature). The high treewidth elsewhere is
a small core: removing $2.95\%$ of mf terms leaves $97\%$ exact, $7.7\%$ of bp leaves $92\%$
exact. Hence $92$--$100\%$ of GO admits \emph{modular} exact WMC---each low-treewidth module
an SDD with weighted separator messages, approximation confined to the small core
(\S\ref{sec:scaling}).

\item \textbf{A full NeSy-framework comparison.}
We run the canonical frameworks on the identical GO inference (validated to reproduce our
exact marginals). Exact compilation (DeepProbLog) walls at ${\sim}400$ terms, one to two
orders below GO's $2.8$k/$6.9$k/$21$k; top-$k$ provenance (Scallop) scales to all $21{,}324$
bp terms in $0.9$\,s but drifts exactly at our reconvergences and is unverifiable at scale;
learned approximators (A-NeSI, NeSyDM) collapse on coupled structure ($0.97\!\to\!0.23$ and
$0.96\!\to\!0.03$--$0.13$ from $N{=}1$ to $N{=}4$). Each trades correctness for scale; ours
is uniquely both scalable and accuracy-characterized (\S\ref{sec:nesy}).
\end{enumerate}

Together these turn the soft closure from an uncharacterized relaxation into an inference
layer with a proof: machine-checked exact on the tree fragment, modular-exact over
$92$--$100\%$ of a real ontology, and bounded-approximate only on a small, identified
reconvergent core. A companion paper applies it to protein function prediction, giving any
base predictor training-free, similarity-independent reach over ontology structure.

\section{Preliminaries}\label{sec:prelim}

\paragraph{Weighted model counting and its $\#$P-hardness.}
Let $\mathcal{Y}=\{1,\dots,m\}$ be a target vocabulary and $T$ a propositional theory over it. Given per-variable weights $\bC p\in[0,1]^m$ factorizing a product distribution $q(\bC y)=\prod_i p_i^{y_i}(1-p_i)^{1-y_i}$, the \emph{weighted model count} of $\phi$ is $\WMC(\phi;\bC p)=\sum_{\bC y\models\phi}\prod_i p_i^{y_i}(1-p_i)^{1-y_i}$. Conditioning $q$ on $T$ yields the structure-aware posterior $p_T(\bC y)\propto[\bC y\models T]\,q(\bC y)$, whose per-target marginals are ratios of weighted model counts, $\mu_i=p_T(y_i{=}1)=\WMC(T\wedge y_i;\bC p)/\WMC(T;\bC p)$---exactly the object underlying semantic loss and the semantic probabilistic layer \citep{xu2018semanticloss,ahmed2022spl}. Weighted model counting subsumes model counting, hence is $\#$P-hard \citep{chavira2008wmc}: naive summation over $2^m$ assignments is hopeless at ontology scale.

\paragraph{Knowledge compilation and treewidth.}
The standard route to tractable $\WMC$ is \emph{knowledge compilation}: transform $T$ once into a circuit (d-DNNF, or the more canonical sentential decision diagram, SDD) on which weighted counting is a single arithmetic pass \citep{darwiche2002kcmap,darwiche2002factoring,darwiche2011sdd}. The $\#$P-hardness is then paid once at compile time, but the compiled circuit's size---and so compilation and evaluation cost---is $\Theta(m\,2^{\tw})$ in the \emph{treewidth} $\tw$ of the constraint graph \citep{darwiche2011sdd,bodlaender1993treewidth}. Treewidth measures how far the graph departs from a tree; it is the single quantity deciding whether monolithic exact $\WMC$ is feasible, and we measure it directly on the ontology graph.

\paragraph{$\mathcal{EL}^{++}$, the true-path rule, and the ELK completion.}
Our symbolic theories are propositional groundings of $\mathcal{EL}^{++}$ ontologies (the OWL 2 EL profile) \citep{baader2005elenvelope}. $\mathcal{EL}^{++}$ is Horn, with a single canonical model and polynomial classification, which is what makes the associated $\WMC$ Boolean and tractable in structure. Reasoning normalizes axioms to the forms $A\sqsubseteq B$, $A\sqcap B\sqsubseteq E$, $A\sqsubseteq \exists r.B$, $\exists r.B\sqsubseteq E$ and saturates them; the ELK reasoner is the canonical realization \citep{kazakov2014elk}. The hierarchical backbone of the Gene Ontology (GO) \citep{ashburner2000go,geneontology2023} is the subsumption (\textit{is-a}/\textit{part-of}) edge set, whose grounding is the \emph{true-path rule}: when a child holds, every subsumption ancestor holds. The resulting \emph{true-path theory} (Definition~1, \S\ref{sec:setup}) is the subsumption edges plus the $\mathcal{EL}^{++}$ normal-form definitions. NeSy predictors over GO do not compute its $\WMC$ marginals $\mu_i$; they apply the monotone message-passing fixpoint of the completion---each node aggregating a normalized soft-OR of its sufficient conditions, capped by the true-path constraint---as a soft relaxation. When this \emph{soft closure} coincides with exact $\WMC$ is the paper's central question.

\paragraph{Belief propagation and its exactness on trees.}
Belief propagation (BP) computes marginals by combining incoming messages as if independent \citep{pearl1988probabilistic}, and is \emph{exact on trees}: on a forest that independence is correct. On cyclic graphs the same updates---\emph{loopy} BP---only approximate. The obstruction is not a two-parent node but a \emph{reconvergence}: two paths meeting at a common descendant, double-counting a shared ancestor. Reading the soft closure as BP turns ``the relaxation is approximate'' into a sharp claim: BP is exact iff the constraint graph is a polytree, tying the error directly to the treewidth and reconvergence statistics of the ontology graph.

\paragraph{The neuro-symbolic landscape.}
NeSy learning couples neural perception with symbolic reasoning, almost universally under the assumption that the marginals $p_i$ may be treated as an independent product \citep{deraedt2020survey,vankrieken2024independence}. Frameworks differ in how they evaluate the constraint, spanning exact-but-costly to scalable-but-uncharacterized. \emph{DeepProbLog} evaluates neural-predicate queries by exact knowledge compilation \citep{deraedt2007problog,manhaeve2018deepproblog}, inheriting ProbLog's exactness and $\#$P-bounded cost. \emph{Scallop} replaces exact provenance with \emph{top-$k$} proofs \citep{huang2021scallop,li2023scallop}, scaling to the full ontology but truncating proofs so its drift concentrates at reconvergent (multi-proof) nodes. \emph{Semantic loss}/\emph{probabilistic layers} compile the constraint and train through it \citep{xu2018semanticloss,ahmed2022spl}---the marginals $\mu_i$ we adopt as gold standard. Finally, \emph{A-NeSI} amortizes inference with a neural surrogate trained on $\WMC$ outputs \citep{vankrieken2023anesi}, and \emph{NeSyDM} models the symbolic joint by masked discrete diffusion \citep{vankrieken2025nesydm}; both collapse on coupled structure (\S\ref{sec:nesy}). Across the spectrum each method trades correctness for scale on the coupled-structure regime---the gap the paper closes.

\section{The symbolic theory and its soft closure}\label{sec:setup}

A neuro-symbolic predictor over an ontology has two components meeting at a probabilistic
interface: a perception model scoring each concept against the input, and a symbolic layer
enforcing the ontology's theory. Under the near-universal independence assumption, the perception
model reports one marginal per concept and the symbolic layer consumes them as if independent. We
develop in turn the four objects this rests on---the factorized perception distribution, the
logical theory, the exact posterior, and the soft relaxation actually deployed.

\paragraph{Independent perception.}
Let $\mathcal{Y}=\{1,\dots,m\}$ be the target vocabulary (for protein function prediction, the $m$
GO terms) and $f_\theta$ a perception model emitting independent per-concept marginals
$p_i\in[0,1]$. These induce the factorized distribution
$q(\bC y)=\prod_{i=1}^{m} p_i^{y_i}(1-p_i)^{1-y_i}$
over $\bC y\in\{0,1\}^m$. This product form is the operational content of the independence
assumption: $q$ places mass on all $2^m$ configurations, including the majority the ontology
forbids. The symbolic layer removes that mass and renormalizes, coupling the scores through the
logical structure.

\paragraph{The true-path theory.}
The relevant logical structure is the propositional theory generated by the ontology's
hierarchical and definitional axioms, restated as a definition.

\begin{definition}[True-path theory]
A theory $T$ is a propositional formula over $\mathcal Y$. The \emph{true-path} theory is
\[
T=\bigwedge_{(c,\pi)\in E}\bigl(y_c\!\Rightarrow\!y_\pi\bigr)\;\wedge\;
\bigwedge_{A\sqcap B\sqsubseteq E}\bigl(y_A\wedge y_B\!\Rightarrow\!y_E\bigr)\;\wedge\;(\text{existential defs}):
\]
the hierarchy edges $E$ plus the EL\textsuperscript{++} normal-form definitions
\citep{baader2005elenvelope,kazakov2014elk}.
\end{definition}

The first conjunction is the \emph{true-path rule} of the GO \citep{ashburner2000go}: when a term
$c$ holds, every more general $\pi$ above it on a subsumption edge holds too. The remaining
conjuncts capture the EL\textsuperscript{++} normal forms a real ontology compiles to
\citep{baader2005elenvelope,kazakov2014elk}, so $T$ is the grounding of the full Horn fragment,
not merely a tree of edges. Being Horn, $T$ has a single canonical (least) model relative to any
asserted facts, making its closure a deterministic upward propagation rather than a branching
search.

\paragraph{The exact posterior: a WMC marginal.}
Conditioning $q$ on the theory gives the principled posterior, whose per-concept marginals are
ratios of weighted model counts.

\begin{definition}[WMC marginal]
The structure-conditioned posterior is $p_T(\bC y)\propto[\bC y\models T]\,q(\bC y)$, with target
marginal
\[
\mu_i=p_T(y_i{=}1)=\frac{\WMC(T\wedge y_i;\bC p)}{\WMC(T;\bC p)},\qquad
\WMC(\phi;\bC p)=\!\!\sum_{\bC y\models\phi}\prod_i p_i^{y_i}(1-p_i)^{1-y_i}.
\]
\end{definition}

Here $[\,\cdot\,]$ is the Iverson bracket: summing only over models of $\phi$ restricts the count
to admissible configurations, so $\mu_i$ is the probability concept $i$ holds once impossible
assignments are excluded and mass renormalized. This is exactly the semantic-probabilistic-layer
construction \citep{ahmed2022spl,xu2018semanticloss}. Evaluating it exactly proceeds by knowledge
compilation of $T$ to an SDD or d-DNNF \citep{darwiche2002kcmap,darwiche2011sdd}; the catch is
cost: $\WMC$ is $\#$P-hard \citep{chavira2008wmc} and the circuit size is governed by treewidth
(\S\ref{sec:scaling}), prohibitive on the largest GO namespace.

\paragraph{The soft closure.}
Because exact compilation does not scale, the inference actually deployed is not the WMC marginal
but the monotone message-passing fixpoint of the EL completion \citep{kazakov2014elk}: each node
aggregates a normalized soft-OR of its sufficient conditions, capped by the true-path constraint
linking it to its children. A parent combines incoming child messages \emph{as if independent},
mirroring at the probability level the deterministic upward propagation the Horn closure performs
at the truth-value level. Iterated to convergence, this is precisely loopy belief propagation
\citep{pearl1988probabilistic} on the constraint graph of $T$. The fixpoint always exists (the
closure is extensive and monotone, reaching its limit within the graph's depth), but \emph{what}
it converges to is structure-dependent, since merging messages as independent is justified only
when the supporting derivations share no evidence. Locating where this is exact versus where it
drifts is the rest of the paper: machine-checked exact on the tree fragment for all priors with
error confined to reconvergences (\S\ref{sec:lean}), made tractable at scale by a modular
construction (\S\ref{sec:scaling}), and positioned against the NeSy spectrum (\S\ref{sec:nesy}).

\section{The soft closure is belief propagation: machine-checked exactness}\label{sec:lean}

The soft closure of \S\ref{sec:setup} is the inference primitive almost every NeSy system runs under the independence assumption \citep{kazakov2014elk,pearl1988probabilistic}. Because it is what is deployed---rather than the \#P-hard exact WMC it stands in for \citep{chavira2008wmc}---the central question is not whether exact WMC is correct (it is, by construction) but \emph{when the cheap closure coincides with it, and where it cannot}. A deployable layer needs this as a theorem. We therefore formalize the closure and its relation to exact WMC in Lean~4 \citep{demoura2021lean}, kernel-checked, tracking the axiom footprint throughout.

The formalization is split into a \emph{finite, concrete} development in Lean core (worlds as Boolean lists, the true-path constraint as edge clauses, \texttt{wmcMarg} as a ratio of summed world weights, every claim closed by kernel \texttt{decide} over a custom cross-multiplied rational encoding) and a \emph{parametric} \texttt{mathlib} development quantified over all $k$ and all real priors. The concrete development's axioms stay within $\{\texttt{propext},\texttt{Classical.choice},\texttt{Quot.sound}\}$ (the \texttt{WMCBattery} within $\{\texttt{propext}\}$ alone); no \texttt{sorry} appears anywhere. The implementation---kernel-reducible rational arithmetic, the proof-structure telescoping, and the full lemma inventory---is in Appendix~\ref{app:lean}.

\paragraph{Tree-exactness for all priors.}
The inductive primitive is the \emph{star}: a target $\pi$ with $k$ children $c_1,\dots,c_k$ independent under the single true-path constraint $y_{c_j}\!\Rightarrow\!y_\pi$. This is the elementary tree fragment, where the soft-OR update forces $\pi$ true unless every child is absent.

\begin{theorem}[Tree-exactness, all priors]\label{thm:star}
For a target $\pi$ with $k$ independent children $c_1,\dots,c_k$ under the true-path constraint $y_{c_j}\!\Rightarrow\!y_\pi$, the exact WMC marginal of $\pi$ equals the soft-OR update
\[
\mu_\pi=\frac{q_\pi}{q_\pi+(1-q_\pi)\prod_{j=1}^{k}(1-q_{c_j})}
\]
for all $k\in\mathbb{N}$ and all priors $q_\pi,q_{c_1},\dots,q_{c_k}$ (Lean: \texttt{star\_marginal\_eq\_softOR}).
\end{theorem}

This is the \texttt{mathlib} theorem \texttt{WmcGeneral.star\_marginal\_eq\_softOR}, stated over $k:\mathbb{N}$ children indexed by $\mathrm{Fin}\,k$ and real priors, hence genuinely universal rather than a finite grid. The proof telescopes the numerator's product-of-sums over all $2^k$ child assignments to $q_\pi$ and adds the single parent-false world contributing $(1-q_\pi)\prod_i(1-q_i)$, so the marginal is the soft-OR ratio for every $k$ and prior (proof detail in Appendix~\ref{app:lean}). Theorem~\ref{thm:star} is thus the machine-checked instance, for the true-path theory, of the classical fact that BP is exact on trees \citep{pearl1988probabilistic}: combining messages as independent is justified precisely when the children are independent.

\paragraph{The error is confined to reconvergences.}
\emph{Where} does the closure deviate from WMC, and \emph{by how much}? The closure equals exact inference on the \emph{computation-tree unrolling}---duplicating any node reached along more than one path---so it is exact iff the constraint graph is a polytree. The obstruction is not a two-parent node but a \emph{reconvergence}, where two paths meet at a shared ancestor. The canonical witness is the diamond $d\!\Rightarrow\!b,\,d\!\Rightarrow\!c,\,b\!\Rightarrow\!a,\,c\!\Rightarrow\!a$. At the uniform prior $\tfrac12$, the exact WMC marginal at the reconvergent node $a$ is $5/6$, whereas the closure (WMC on the tree splitting $a$ into $a_1,a_2$) gives $7/10$ (Lean: \texttt{diamond\_reconv\_inexact}, with values pinned by \texttt{diamond\_exact\_a\_is\_5\_6} and \texttt{diamond\_bp\_a\_is\_7\_10}); at the sink the exact $1/6$ disagrees with the BP form $1/5$ (\texttt{diamond\_sink\_bp\_wrong}), while the polytree sink is exact at $1/5$ (\texttt{polytree\_sink\_exact}). The kernel-checked breadth battery \texttt{WMCBattery.lean} confirms over a $9\times9=81$-prior grid that the single edge is exact (\texttt{single\_edge\_exact\_all\_priors}), the polytree sink is exact (\texttt{polytree\_exact\_all\_priors}), and the diamond is \emph{wrong} (\texttt{diamond\_wrong\_all\_priors}) across all priors---so exactness is a property of acyclicity-up-to-reconvergence, not of the prior. This is the quantity the GO structure probe measures (\S\ref{sec:scaling}) and the cut criterion for the modular construction; per-prior numbers are tabulated in Appendix~\ref{app:lean}.

\paragraph{Unconditional convergence of the monotone fixpoint.}
The synchronous true-path step is \emph{extensive} (\texttt{step\_ext}) and \emph{monotone} (\texttt{step\_mono}); in a finite vocabulary the active set grows monotonically and reaches the least fixpoint in at most $\mathrm{depth}$ steps, fixed once a step adds nothing (\texttt{fixed\_stable}, requiring \emph{no axioms}). Convergence is thus unconditional; only the limit is structure-dependent. What lies outside the machine-checked perimeter is a closed-form error bound on general DAGs (\#P-hard), reported empirically against the modular-exact SDD reference in \S\ref{sec:scaling}.

\section{Tractability: treewidth and modular exact WMC}\label{sec:scaling}
Theorem~\ref{thm:star} establishes \emph{when} the soft closure is correct; a deployable layer must also answer \emph{at what cost} the exact answer is recoverable off the tree, and \emph{how much} of a real ontology falls outside the tree regime. We make both concrete for the Gene Ontology (GO) and turn them into a constructive recipe: a modular exact-WMC scheme whose only inexactness is the bounded soft approximation of Theorem~\ref{thm:star}, confined to a small, identified reconvergent core.

\paragraph{Exact WMC is governed by treewidth.}
Exact $\mu_i$ proceeds by compiling $T$ to an SDD, equivalently by junction-tree calibration, at cost $\Theta(m\,2^{\tw})$ in the constraint-graph treewidth $\tw$ \citep{darwiche2011sdd,darwiche2002kcmap,bodlaender1993treewidth}, $m=|\mathcal Y|$. On a forest $\tw=1$ and exact WMC coincides with the single-pass BP of Theorem~\ref{thm:star}; each reconvergence enlarges the largest junction-tree bag. The exponential dependence on $\tw$ is exactly why a soft closure is used, so quantifying $\tw$ on the actual ontology is the pivot of the tractability argument.

\paragraph{Measured GO treewidth.}
We compute a min-fill treewidth of the GO true-path graph (built from the EL\textsuperscript{++} normal forms of \texttt{go.norm}, one clique per axiom factor) per namespace:
\[
\tw = 9~(\text{cellular component, cc}),\quad
31~(\text{molecular function, mf}),\quad
272~(\text{biological process, bp}).
\]
$2^{9}$ is trivial; $2^{31}\!\approx\!2\times10^{9}$ is borderline-infeasible as a monolith; and bp's single component at $\tw=272$ would need a circuit of size ${\sim}2^{272}$, out of reach. Treewidth is a \emph{graph invariant}, so no decomposition computes exact WMC below it: bp admits \emph{no} exact monolithic WMC.

\paragraph{cellular component: exact WMC validates the localized error.}
On cc we min-fill-decompose and run two-pass Shafer--Shenoy calibration for exact per-term marginals: $2{,}829$ terms, three components, max treewidth $9$, full exact marginals in $7.5$\,s. Against this reference the soft closure agrees to mean error $6\times10^{-4}$ (p95 $1.3\times10^{-3}$, max $0.076$), only $2.0\%$ of terms off by ${>}0.01$. The error is not uniform: mean error at multi-parent (reconvergent) terms is $2.1\times10^{-3}$ versus $3\times10^{-4}$ at single-parent terms---roughly $7\times$ larger, exactly the reconvergence signature Theorem~\ref{thm:star} predicts. The closure is near-exact precisely because cc is near-tree-like, and the residual localizes onto the diamond reconvergences (\texttt{diamond\_reconv\_inexact}) the Lean development isolates.

\paragraph{Cutting a small reconvergence core restores exactness.}
The high treewidth of mf and bp is not diffuse but carried by a few high-degree reconvergence hubs (max nf1 in-degree $4/7/8$ on cc/mf/bp). Greedily deleting the highest-degree terms in $1\%$ batches and recomputing min-fill until the remainder is exact ($\tw\le14$) yields a small core: on mf, cutting $202$ terms ($2.95\%$) leaves $97.05\%$ exactly WMC-computable; on bp (worst case, starting from $\tw\approx269$), cutting $1648$ terms ($7.7\%$) leaves $92.28\%$ exactly computable. So a core of at most $7.7\%$ of terms accounts for essentially all intractability, and the complementary $92$--$100\%$ of GO becomes exactly computable once it is separated. bp---$64\%$ multi-parent, circuit rank $24{,}432$---is exactly where a small, unavoidable core must be handled by bounded approximation; the measurement shows that core is small and identifiable, not pervasive.

\paragraph{A modular construction.}
The low-treewidth periphery and small high-treewidth core combine into a scheme recovering exact WMC over almost all of GO while confining approximation to the characterized regime.

\begin{proposition}[Modular exact WMC]\label{prop:modular}
Partition the constraint graph into low-treewidth modules separated by the reconvergence core. Compile each module to an SDD, evaluate its exact WMC, and pass weighted separator marginals between modules; the result is exact on the union of modules, with the bounded soft approximation (Theorem~\ref{thm:star}) confined to the small core.
\end{proposition}

The construction generalizes the cc junction-tree calibration: each module's exact SDD is cheap, and inter-module coordination passes weighted marginals over the low-treewidth separators rather than Boolean entailments. Where a module interface is a tree the messages are exact; only across the residual ${\le}8\%$ core do they fall back on the soft closure, bounded by Theorem~\ref{thm:star}. The net effect is exact WMC over $92$--$100\%$ of GO with a single small locus of bounded approximation---unlike a uniform soft closure or a monolithic compilation that cannot be built for bp.

\section{Comparison to neuro-symbolic frameworks}\label{sec:nesy}

Having characterized the closure (\S\ref{sec:lean}) and made it tractable (\S\ref{sec:scaling}), we
now ask where the canonical NeSy frameworks sit on the identical problem. The space is organized by
how each attacks the $\#$P-hardness of $\WMC$ \citep{chavira2008wmc,deraedt2020survey}: exact
knowledge compilation keeps correctness at full circuit cost; provenance relaxations truncate proof
structure and trade correctness for scale; learned approximators amortize or model the joint and
trade correctness for a training-bound notion of scale. We run a representative of each family on the
\emph{identical} GO true-path inference and read off which property each keeps. The positioning is
summarized in Table~\ref{tab:nesy}; ours is the only entry that keeps both columns.

\begin{table}[t]\centering
\caption{Neuro-symbolic inference on the same GO true-path WMC.}\label{tab:nesy}
\small
\begin{tabular}{lcc}
\toprule
method & scales to full GO? & accuracy at scale\\
\midrule
DeepProbLog \citep{manhaeve2018deepproblog} (exact KC) & no (wall ${\sim}400$ terms) & exact when it runs\\
Scallop \citep{huang2021scallop} (top-$k$) & yes (${<}1$\,s) & no guarantee; drift at reconvergence\\
A-NeSI \citep{vankrieken2023anesi} / NeSyDM \citep{vankrieken2025nesydm} & learned & collapse on coupled structure ($N{=}4$)\\
\textbf{ours} (soft-WMC $+$ modular) & yes & exact on tree (Lean) $+$ modular-exact $92$--$100\%$\\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Setup: a like-for-like inference, validated against our exact marginals.}
Every system must compute \emph{the same} structure-conditioned marginals $\mu_i$ of $T$ under the
factorized prior $\bC q$, so we encode $T$ natively in each (hierarchy edges as ProbLog rules with
the true-path constraint as evidence for DeepProbLog; the upward derivation
$\texttt{holds}(P)\!:\!-\,\texttt{sub}(C,P),\texttt{holds}(C)$ for Scallop). Both tools reproduce our
exact answers on validation cases: DeepProbLog's ProbLog engine returns
$P(\texttt{go}(a))=0.8333=5/6$ and $P(\texttt{go}(d))=1/6$ on the diamond (Theorem~\ref{thm:star}'s
reconvergence instance), and Scallop returns $P(\texttt{holds}(x))=0.875$ on the two-parent chain at
$p=\tfrac12$. So any divergence at scale is a property of the inference strategy, not the encoding.

\paragraph{Exact knowledge compilation walls one to two orders below GO scale.}
DeepProbLog compiles the grounded program to a $\#$P circuit and reads off marginals
\citep{manhaeve2018deepproblog,chavira2008wmc}---the same exact route our monolithic SDD baseline
takes---so its cost is governed by treewidth (\S\ref{sec:scaling}). Sweeping subgraph size on mf/bp,
it compiles $100$ terms in $0.3$\,s and $250$ in $0.8$--$2.0$\,s, then walls at ${\sim}400$ terms
($>300$\,s timeout). The full namespaces---$2.8$k/$6.9$k/$21$k terms---are one to two orders beyond
that ceiling. The wall is rigorous, not incidental: bp's treewidth $272$ implies a $2^{272}$ circuit
no exact compiler can build. Exact NeSy is correct but cannot reach ontology scale.

\paragraph{Top-$k$ provenance scales, but its drift lives exactly at the reconvergences our theory identifies.}
Scallop computes differentiable provenance over only the $k$ highest-weight proofs per fact
\citep{huang2021scallop,li2023scallop}, evaluating the entire $21{,}324$-term bp namespace in
$0.9$\,s. The cost is correctness, and the error appears where Theorem~\ref{thm:star} predicts:
top-$k$ loses mass only when a fact has more than $k$ derivations, i.e.\ at reconvergent (multi-proof)
nodes, and is already exact on the tree fragment. A four-node reconvergent derivation (exact marginal
$0.9375$) is undercounted as $0.50$ at $k{=}1$, recovering to $0.875$ at $k{=}3$ and exactly $0.9375$
at $k{=}10$; on $14$-node bp neighborhoods the max drift falls from $0.15$--$0.30$ at $k{=}1$ to
$0.000$ at $k{=}10$ (Appendix~\ref{app:nesy}). The decisive point is at scale: a hub term (rank
$24{,}432$) has astronomically many proofs, so any fixed $k$ undercounts, and that undercount is
\emph{unverifiable}---the $\#$P-hardness made concrete. Scallop scales by giving up the guarantee,
exactly where our theory locates the difficulty.

\paragraph{Learned approximators relocate the wall to coupled structure.}
A-NeSI \citep{vankrieken2023anesi} and NeSyDM \citep{vankrieken2025nesydm} learn $\WMC$ rather than
compute it---A-NeSI amortizing the output with a neural surrogate, NeSyDM modelling the joint by
masked discrete diffusion. They are training methods, not inference engines, hence absent from the
scaling sweep: deploying on GO means a fresh model per ontology, and A-NeSI's surrogate must moreover
train on \emph{samples of the exact $\WMC$}---intractable to generate at GO scale, the very
$\#$P-hardness we attack. Our layer instead computes the exact tree-fragment marginal with no
inference-training data. On the coupled-structure carry-chain benchmark both collapse as depth grows:
A-NeSI degrades from $0.968$ at $N{=}1$ to $0.225$ at $N{=}4$, and NeSyDM from $0.962$ to
$0.03$--$0.13$ (three seeds; per-seed numbers in Appendix~\ref{app:nesy})---the
independence/reasoning-shortcut failure \citep{vankrieken2024independence,marconato2023reasoningshortcuts}
this project targets. The learned route relocates the wall from exact-inference cost to approximation
quality on coupled structure, again without a guarantee.

\paragraph{Positioning.}
Each family---exact compilation, top-$k$ provenance, amortized inference, joint diffusion---trades
correctness for scale, with no guarantee on the coupled-structure regime: DeepProbLog loses scale at
${\sim}400$ terms; Scallop loses the guarantee at the reconvergences of Theorem~\ref{thm:star}; A-NeSI
and NeSyDM collapse on coupled depth ($N{=}1\!\to\!4$). Our construction is the only entry of
Table~\ref{tab:nesy} in both columns: machine-checked exact on the tree fragment for all priors,
modular-exact over $92$--$100\%$ of GO, and bounded-approximate only on the small reconvergent core
the same theory isolates. Scalability and a correctness characterization are jointly attainable for
NeSy inference over a real, ontology-scale theory.

\section{Application, discussion, and conclusion}

\subsection{Application: training-free, similarity-independent ontology reach}
The layer is a drop-in posterior over the true-path theory: it maps the independent marginals
$p_i$ of \emph{any} perception model to the structure-conditioned $\mu_i$, with no retraining and
no parameters of its own. This makes it directly useful for protein function prediction over the
GO \citep{ashburner2000go,geneontology2023}. Modern sequence-based predictors---convolutional and
transformer annotators \citep{kulmanov2020deepgoplus,bileschi2022protenn,cao2021tale}, protein
language models \citep{rives2021esm1b,lin2023esm2}, contrastive enzyme classifiers
\citep{yu2023clean}---emit per-term scores treated as conditionally independent, with ontology
consistency imposed only heuristically or recovered with supervision
\citep{kulmanov2018deepgo,kulmanov2022deepgozero}. Our layer instead conditions those scores on
the GO is-a/part-of and EL\textsuperscript{++} structure exactly (tree fragment) and
modular-exactly elsewhere, giving reach over ontology structure that is \emph{training-free} and,
because the correction comes from the theory rather than labelled neighbours, \emph{independent of
sequence similarity}. A companion paper develops this in full on standard benchmarks
\citep{radivojac2013cafa,zhou2019cafa3} across several base predictors.

\subsection{Discussion}
The central message is that the soft closure NeSy systems run under the independence assumption
\citep{vankrieken2024independence} is not a black box: for the true-path theory it is loopy belief
propagation \citep{pearl1988probabilistic}, established by a machine-checked guarantee
(Theorem~\ref{thm:star}, no \texttt{sorry}, axioms
$\subseteq\{\texttt{propext},\texttt{Classical.choice},\texttt{Quot.sound}\}$) rather than an
analogy. The same formalization isolates the \emph{single} failure mode---reconvergences---and
quantifies it (\texttt{diamond\_reconv\_inexact}: $5/6$ vs $7/10$). Crucially this predicted mode
carries the empirical error on a real ontology: on cc the soft closure agrees with exact
junction-tree WMC to mean error $6\times10^{-4}$, concentrated $7\times$ at multi-parent terms,
exactly the reconvergence signature. The same fingerprint reappears across the spectrum (Scallop's
top-$k$ drift at multi-proof nodes; the treewidth core is itself reconvergence structure), so
theory, exact computation, and independent NeSy engines agree on \emph{where} the difficulty
lives. The payoff is tractability that would otherwise look impossible: since exact WMC costs
$\Theta(m\,2^{\tw})$ and treewidth is a graph invariant, bp admits \emph{no} exact monolithic WMC,
yet the high treewidth is a small core, so Proposition~\ref{prop:modular} recovers exact WMC over
$92$--$100\%$ of GO with approximation confined to that core---the honest form of ``modules buy
tractability''.

\subsection{Future work}
Two extensions follow. First, the modular construction relies on a generic tree-decomposition;
replacing it with a logically principled one---the consequence-based context calculus
\citep{bate2016sriq,tenacucala2019sequoia} with locality-based modules
\citep{cuencagrau2008modular,delvescovo2011atomic}, lifting messages from \emph{Boolean}
consequences to \emph{weighted} separator marginals---would yield a fully verified modular WMC; the
engine already passes messages over these interfaces, so only the payload changes. Second, the
EL\textsuperscript{++} compilation \citep{baader2005elenvelope,kazakov2014elk} should extend toward
OWL~2~DL via hypertableau reasoning for SROIQ
\citep{horrocks2006sroiq,motik2009hypertableau}, bringing disjunction, number restrictions,
nominals, and role chains within reach while preserving the correctness-characterized core.

\subsection{Conclusion}
The soft closure NeSy systems run under the independence assumption is, for the true-path theory,
belief propagation, and we give it the two properties a deployable layer needs: a machine-checked
correctness characterization (exact on the tree fragment for all priors, error confined to
reconvergences) and tractability (modular-exact over $92$--$100\%$ of GO, approximation bounded to
a small identified core). The layer is uniquely both scalable and accuracy-characterized across the
NeSy spectrum, and directly gives any protein function predictor training-free,
similarity-independent reach over ontology structure.

\paragraph{Data and code availability.}
The Lean~4 developments---including \texttt{star\_marginal\_eq\_softOR},
\texttt{diamond\_reconv\_inexact}, and the kernel-checked \texttt{WMCBattery.lean}---together with
the treewidth and irreducible-core probes, the junction-tree exact-WMC reference, the modular
SDD construction, and the EL\textsuperscript{++} reasoner core, will be released in a public
repository upon publication; an anonymized archive is provided to reviewers.

\acks{Acknowledgements omitted for double-blind review.}

\bibliography{references}

\appendix

\section{Lean formalization: implementation and lemma inventory}\label{app:lean}

\paragraph{Two developments and a kernel-checkable rational arithmetic.}
The formalization is split deliberately. A \emph{finite, concrete} development lives in Lean core with no dependence on \texttt{mathlib}: it defines worlds as Boolean lists, the true-path constraint as a conjunction of edge clauses $(\neg y_c \vee y_\pi)$, the exact weighted model count \texttt{wmcMarg} as a ratio of summed world weights, and the belief-propagation update forms, and it closes every claim by \texttt{decide}---reduction in the kernel, with no \texttt{native\_decide} escape hatch and therefore no trust placed in compiled code. The cost of staying inside the kernel is that ordinary rational arithmetic does not reduce cheaply, so we introduce a \emph{custom rational kernel encoding}: a rational is a pair $Q=\mathbb{Z}\times\mathbb{Z}$ of numerator and denominator, addition, multiplication and complementation ($1-a$) are defined directly on the pair, and equality is the \emph{cross-multiplied} test $a_1 b_2 = b_1 a_2$, decidable on integers. Because equality of $Q$-values never normalizes a fraction and reduces to a single integer identity, the entire finite battery---world enumeration, weighting, summation, and the BP/WMC comparison---is reducible by \texttt{decide} and hence checked in the kernel. The price is that the concrete theorems hold only at the specific priors enumerated; the \emph{parametric} statement, quantified over all $k$ and all real priors, is proved separately in a \texttt{mathlib} development where real-number reasoning and finite-product lemmas are available. The concrete development's axioms stay within $\{\texttt{propext},\texttt{Classical.choice},\texttt{Quot.sound}\}$ (the \texttt{WMCBattery} battery within $\{\texttt{propext}\}$ alone); no \texttt{sorry} appears anywhere.

\paragraph{Proof of Theorem~\ref{thm:star} (\texttt{WmcGeneral.star\_marginal\_eq\_softOR}).}
The statement is over $k:\mathbb{N}$ children indexed by $\mathrm{Fin}\,k$ and priors in $\mathbb{R}$, so it is genuinely universal rather than a finite witness grid. The weight of a world $(p,c)$ factorizes as the parent factor times the product $\prod_i(\text{if }c_i\text{ then }q_i\text{ else }1-q_i)$ over children. The numerator---the unnormalized mass of valid worlds with $\pi$ true---is a sum over \emph{all} $2^k$ children assignments, since when the parent is true every child assignment is consistent with the constraint; this Finset product-of-sums expansion telescopes, because the sum over each child of (true-weight $+$ false-weight) is $1$ (\texttt{sum\_children}, via $\prod_i\sum_b = \sum_{c}\prod_i$ on \texttt{Fintype.piFinset}), giving $\mathrm{num}=q_\pi$ exactly (\texttt{num\_eq}). The partition function adds the \emph{single} valid world in which the parent is false---which forces all children false---contributing $(1-q_\pi)\prod_i(1-q_i)$, so $Z=q_\pi+(1-q_\pi)\prod_i(1-q_i)$ (\texttt{Zp\_eq}). The marginal is their ratio, exactly the soft-OR form, for every $k$ and every prior.

\paragraph{Reconvergence witnesses and the breadth battery.}
The closure equals exact inference on the \emph{computation-tree unrolling} of the constraint graph---the tree obtained by duplicating any node reached along more than one path---so it is exact if and only if the unrolling is the identity, i.e.\ if and only if the constraint graph is a polytree. On the diamond at uniform prior $\tfrac12$, the reconvergent node $a$ has exact value $5/6$ (\texttt{diamond\_exact\_a\_is\_5\_6}) versus closure $7/10$ (\texttt{diamond\_bp\_a\_is\_7\_10}); the sink's exact $1/6$ disagrees with the BP form $1/5$ (\texttt{diamond\_sink\_bp\_wrong}), while the polytree sink is exact at $1/5$ (\texttt{polytree\_sink\_exact}). The breadth battery \texttt{WMCBattery.lean} evaluates the same comparisons over nine rational priors in $(0,1)$, all kernel-checked by \texttt{decide}: the single-edge BP update equals the exact WMC parent marginal across the full $9\times 9=81$-prior grid (\texttt{single\_edge\_exact\_all\_priors}); the independent two-parent polytree sink is exact across all priors (\texttt{polytree\_exact\_all\_priors}); and the diamond's BP independence form is \emph{wrong} across all priors (\texttt{diamond\_wrong\_all\_priors}).

\paragraph{Convergence lemmas.}
The synchronous true-path step---a parent is activated by any active child---is proved \emph{extensive} (it only ever adds active nodes, \texttt{step\_ext}) and \emph{monotone} in the active set (\texttt{step\_mono}). Since the active set grows monotonically inside a finite vocabulary, iteration reaches the least fixpoint in at most $\mathrm{depth}$ steps, and once a step adds nothing every further iteration is fixed (\texttt{fixed\_stable}, which requires \emph{no axioms at all}).

\section{Extended neuro-symbolic comparison numbers}\label{app:nesy}

\paragraph{Top-$k$ provenance (Scallop).}
The four-node reconvergent derivation has two proof paths to a shared fact via the upward rule $\texttt{holds}(P)\!:\!-\,\texttt{sub}(C,P),\texttt{holds}(C)$ at edge-probability $\tfrac12$---a construction distinct from the Lean true-path diamond of \S\ref{sec:lean} (different inference semantics, so its exact value $0.9375$ does not conflict with that diamond's $5/6$). The exact marginal $P(\texttt{holds}(a))=0.9375$ is approximated as $0.50$ at $k{=}1$ (a $47\%$ undercount), recovers to $0.875$ at $k{=}3$, and returns to the exact $0.9375$ at $k{=}10$. On $14$-node biological-process neighborhoods the maximum drift versus the exact reference falls from $0.15$--$0.30$ at $k{=}1$ to $0.000$ at $k{=}10$. The error thus grows with proof multiplicity (reconvergence) and shrinks as $k$ admits more proofs---the same reconvergence axis along which our soft closure departs from exact $\WMC$.

\paragraph{Learned approximators (A-NeSI, NeSyDM).}
On the $N$-digit carry-chain benchmark, measured over three seeds: A-NeSI degrades from $0.968$ at $N{=}1$ to $0.225$ at $N{=}4$, and NeSyDM from $0.962$ at $N{=}1$ to $0.03$--$0.13$ at $N{=}4$. At $N{=}4$ the carry constraint binds four digits and both learned approximators break down---the independence-assumption / reasoning-shortcut failure \citep{vankrieken2024independence,marconato2023reasoningshortcuts} this project targets.

\end{document}
